---
title: "wrangling"
author: "J Coetsee - 19491050"
date: "25/06/2021"
output: pdf_document
---

```{r setup, include=FALSE}
#chunk options
knitr::opts_chunk$set(echo = TRUE)

#source functions from code folder:
list.files('code/', full.names = T, recursive = T) %>% as.list() %>% walk(~source(.))

library(pacman)
p_load("scales","cowplot", "randomForest", "haven", "dbplyr","RSQLite", "tidyverse")
theme_set(hrbrthemes::theme_ipsum())

```

# Read in data to be converted into local sql database

```{r}
#select and write derived data
#der <- read_dta("data/derived_NIDS-CRAM_Wave1_Anon_V2.0.0.dta") 
#
#der <- der %>% 
#    select(pid, w1_nc_best_age_yrs, w1_nc_age_intervals, w1_nc_best_race, #w1_nc_empl_stat, w1_nc_geo2011, w1_nc_best_gen, w1_nc_mdbdc2011)
#
#write.csv(der, "data/derivedwav1.csv")

#read data
data <- read_csv("data/nidswav1.csv") %>% 
    select(-1) 

deriveddata <- read_csv("data/derivedwav1.csv") %>% 
    select(-1) %>% 
    rename(derpid = pid)



```

# Create SQLite database 

```{r}

#SQL db with two tables, one with the main dataset, and one where some variables have been derived by the NIDSCRAM team:

mydb <- DBI::dbConnect(SQLite(), dbname = "")
dbWriteTable(mydb, "nids", data)
dbWriteTable(mydb, "derived", deriveddata)
dbListTables(mydb)
```

# Focus:

Want to compute a RF to look at the effect of Lockdown on whether households have lost their main source of income. i.e. who are most likely to have lost their jobs due to lockdown. 

------------------------------------------------------

# Selecting the relevant variables that have enough observations from nids table:

variable I want to classify/predict:

w1_nc_hhincchng = da10_9 - "Has household lost main source of income since lockdown start 27th Mar"

------------------------------------------------------

# Other Features: (16 in total, excluding w1_nc_hhincchng and pid)

pid = Person identifier

w1_nc_hhinc = da9 - Total household income after tax in April # 2227 reported doesn't know.

w1_nc_hhincdec1 = da8_1 - Sources of household income decreased during lockdown? 1

w1_nc_hhincsrc1 = da7_1 - Sources household income in February? 1

w1_nc_incgov = da5 - Do you receive any kind of government grant?

w1_nc_unemdc = 	cg2 - When was the last time you worked?

w1_nc_emwrk_isco_c = cd1 - Occupational code for usual work

w1_nc_emtyp = cb6 - Respondent's main form of work

w1_nc_enrgelec = b17 - Dwelling/house has access to electricity?

w1_nc_watsrc =  b15 - Piped or tap water inside dwelling/house/in yard?

w1_nc_nocld = b14 - More/less/same number kids in house now compared to before the lockdown?

w1_nc_nopres = b10 - Number of people resident, including youself (don't forget babies)

w1_nc_dwltyp = b9 - Type of dwelling or house living in

w1_nc_moveres_apr = b7 - Moved to another house/dwelling within the province for April lockdown?

w1_nc_prov = b6_1 - Province currently living in now?

w1_nc_edter = b4 - Respondent successfully completed some form of tertiary studies?

w1_nc_edschgrd = b3 - Highest school grade completed

------------------------------------------------------

# and from derived table:

derpid = pid

w1_nc_best_age_yrs = age in years

w1_nc_age_intervals = age intervals (5 years)

w1_nc_best_race = race/population group

w1_nc_empl_stat = employment status

w1_nc_geo2011 = geotype (based on 2011 census, traditional-urban-farms)

w1_nc_best_gen = gender

w1_nc_mdbdc2011 = municipality (2011 census)

------------------------------------------------------

# Making the complete SQL database with all variables:

```{r}

full <- tbl(mydb, sql("SELECT pid, w1_nc_hhincchng, w1_nc_hhinc, w1_nc_hhincdec1, w1_nc_incgov, w1_nc_hhincsrc1, w1_nc_unemdc, w1_nc_emwrk_isco_c, w1_nc_emtyp,w1_nc_enrgelec, w1_nc_watsrc, w1_nc_nocld, w1_nc_nopres, w1_nc_dwltyp, w1_nc_moveres_apr, w1_nc_prov, w1_nc_edter, w1_nc_edschgrd,w1_nc_best_age_yrs, w1_nc_age_intervals,w1_nc_best_race, w1_nc_empl_stat,w1_nc_geo2011, w1_nc_best_gen, w1_nc_mdbdc2011  FROM nids, derived WHERE pid = derpid")) %>% 
    collect()

```

------------------------------------------------------

# CLEANING: Missing Data/NA's

```{r}
# check amount of NA's by column
colSums(is.na(full))

full <- full %>% 
    mutate(w1_nc_hhincchng = ifelse(w1_nc_hhincchng < 0, NA, w1_nc_hhincchng), # hhincome change, change to NA, there are very few values. need no ,missing for this var
           w1_nc_hhinc = ifelse(w1_nc_hhinc < 0, 2800, w1_nc_hhinc), # missing
           w1_nc_hhincdec1 = ifelse(w1_nc_hhincdec1 < 0, -3, w1_nc_hhincdec1), #missing
           w1_nc_incgov = ifelse(w1_nc_incgov < 0, -3, w1_nc_incgov), #missing
           w1_nc_hhincsrc1 = ifelse(w1_nc_hhincsrc1 < 0, -3, w1_nc_hhincsrc1), #missing
           w1_nc_emtyp = ifelse(w1_nc_emtyp < 0, -3, w1_nc_emtyp), #missing
           w1_nc_enrgelec = ifelse(w1_nc_enrgelec < 0, -3, w1_nc_enrgelec),#missing
           w1_nc_watsrc = ifelse(w1_nc_watsrc < 0, -3, w1_nc_watsrc),#missing
           w1_nc_watsrc = ifelse(w1_nc_watsrc < 0, -3, w1_nc_watsrc),#missing
           w1_nc_nocld = ifelse(w1_nc_nocld < 0, -3, w1_nc_nocld),#missing
           w1_nc_nopres = ifelse(w1_nc_nocld < 0, -3, w1_nc_nocld),#missing
           w1_nc_dwltyp = ifelse(w1_nc_dwltyp < 0, -3, w1_nc_dwltyp), #missing
           w1_nc_moveres_apr = ifelse(w1_nc_moveres_apr < 0, -3, w1_nc_moveres_apr),
           w1_nc_prov = ifelse(w1_nc_prov < 0, -3, w1_nc_prov),
           w1_nc_prov = ifelse(w1_nc_prov == 10, -3, w1_nc_prov),
           w1_nc_edter = ifelse(w1_nc_edter < 0, -3, w1_nc_edter),
           w1_nc_geo2011 = ifelse(w1_nc_geo2011 < 0, -3, w1_nc_geo2011),
           w1_nc_empl_stat = ifelse(w1_nc_empl_stat < 0, -3, w1_nc_empl_stat))

# three problematic columns where NAs > 4300: w1_nc_unemdc, w1_nc_emwrk_isco_c, w1_nc_emtyp; and one where = 1689 (w1_nc_edter)

#for w1_nc_emtyp, w1_nc_unemdc and w1_nc_emwrk_isco_c, the NAs are those without jobs for w1_nc_emtyp. employment status and emtyp therefore captures what we need, so we can throw out the other vars. Similarly, edter gives tertiary education. In this, we assume people who have tertiary would say yes, while the rest do not, so NAs are converted to no's, or 0's

full <- full %>% 
    select(-c(w1_nc_unemdc, w1_nc_emwrk_isco_c)) %>% #drop vars with too many NAs, captured by two other vars
    mutate(w1_nc_emtyp = replace_na(w1_nc_emtyp, 0),  # replace nas with 0's, for unemployed 
           w1_nc_edter = replace_na(w1_nc_edter, 2)) %>% # replace tertiary educ na with 2 (no)
    na.omit()

# Changing colnames:
full <- full %>% 
    rename(  "Income Change" = w1_nc_hhincchng,
             "PID" = pid,
             "HH Income Apr" = w1_nc_hhinc,
             "Sources Income Decreased" = w1_nc_hhincdec1,
             "Grant" = w1_nc_incgov,
             "Sources HH Income" = w1_nc_hhincsrc1,
             "Employment Type" = w1_nc_emtyp,
             "Electricity Access" = w1_nc_enrgelec,
             "Water Access" = w1_nc_watsrc,
             "Children Change" = w1_nc_nocld,
             "HH Size " = w1_nc_nopres,
             "Dwelling Type" = w1_nc_dwltyp,
             "Moved" = w1_nc_moveres_apr,
             "Province" = w1_nc_prov,
             "Tertiary" = w1_nc_edter,
             "Education" = w1_nc_edschgrd,
             "Age" = w1_nc_best_age_yrs,
             "Age Interval" = w1_nc_age_intervals,
             "Race" = w1_nc_best_race,
             "Employed" = w1_nc_empl_stat,
             "Geo Type" = w1_nc_geo2011,
             "Gender" = w1_nc_best_gen,
             "District Council" = w1_nc_mdbdc2011)

write.csv(full, "data/full.csv")

```




